Avoidance in permutations

### 2014-11-07 project2/paper/paper.tex
some edits to section 6

\section{Avoidance of permutations of $T_n$}
\label{Tn}
We now impose further restrictions on the set of permutations $S_n$. This new set of permutations, $T_n$ is defined follows.

\begin{definition}
For even $n$, $T_n$ is defined as the set of all permutations $\sigma \in S_n$ for which $1,3,5,\ldots,2n-1$ appear in increasing order, and $2i$ always appears to the right of $2i-1$.
\end{definition}

Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $0$.

@@

\begin{proof}
Observe that since $1,3,5,\ldots,2m-1$ must appear in increasing order in $T_{2m}$, our problem is now reduced to determining the relative order of $2,4,6,\ldots,2m$ with respect to each other, as well as with respect to $1,3,5,\ldots,2m-1$.

Let us first try to insert $2m$ into the sequence $1,3,5,\ldots,2m-1$. Clearly, the way $T_{2m}$ is defined, $2m$ can be inserted into only $1$ slot, the one following $2m-1$. (Here we define a slot as a gap between two existing elements of the sequence, or the gap that follows the last element of the sequence or the gap that precedes the first element of the sequence)

Now, let us try to insert $(2m-2)$ into the sequence. $(2m-2)$ can be inserted into $3$ slots in the sequence -- the one between $2m-3$ and $2m-1$, the one between $2m-1$ and $2m$ or the one following $2m$.

Now, if we try to insert $(2m-4)$ into this incompletely formed sequence, we will see that there are $5$ possible locations into which this number can be inserted. Continuing this for all even numbers upto $2$, we observe that the total number of ways such a sequence can  be created is equal to $1 \cdot 3 \cdot 5 \cdot \ldots \cdot (2m-1)$, as desired.
\end{proof}

Studying avoidance of permutations in $T_n$ seems an interesting exercise. In the following section, we consider the sequence of numbers $t_n(\pi)$, defined as the number of permutations in $T_n$ that avoid the permutation $\pi$. In particular, we consider permutations $\pi \in S_3$.

@@

More interesting is the equivalences between $213$ and $231$, and $312$ and $321$. We will prove the equivalence between $312$ and $321$ below, but before doing so, we first state and prove the following lemmas which will be useful in our proof.

\begin{lemma}
\label{tn_avoidance}
If $a$ and $b$ are two elements that appear in a permutation in $T_n$ such that $a$ is to the left of $b$, then if $a > b$, $b$ must be even.
\end{lemma}

\begin{proof}
For the sake of contradiction, let us assume that $b$ is odd. Then there exist two possibilities that need to be considered -- one, that $a$ is odd, and two, that $a$ is even.

@@

Note that all permutations in $T_n$ must have odd elements monotonically increasing. So if $a$ appears to the left of $b$, and both $a$ and $b$ are odd, then $a$ must be less than $b$ -- a contradiction. Hence, we conclude that $a$ can't be odd.

@@

For all permutations $\sigma$ in $T_n$, every even element $j$ must be to the right of all odd elements less than $j$. From this, we conclude that since $a$ is even, it must be to the right of all odd numbers less than it. But $b$ is an odd number less than $a$ -- we conclude that $a$ can't be even either.
\end{itemize}

Since $a$ can't be odd or even if $b$ is odd, we conclude that $b$ must be even, as desired.
\end{proof}

\begin{lemma}
\label{swapping_lemma}
If $x, y, z$ is a three-element subsequence of a permutation $\pi$ in $T_n$ such that $x > y > z$ or $x > z > y$, then swapping $y$ and $z$ in the permutation $\pi$ yields a permutation $\pi'$ that also belongs to $T_n$.
\end{lemma}

\begin{proof}
From Lemma \ref{tn_avoidance} we see that both $y$ and $z$ must be even. Then there exist two cases that need to be covered.
\begin{itemize}
\item \textbf{Case 1:} $x$ is odd.

If $x$ is odd, then it is clear that the permutation $\pi'$ obtained from swapping $y$ and $z$ is still valid, since both $y$ and $z$ are to the right of the odd numbers $y-1$ and $z-1$ (since $y-1$ and $z-1$ are themselves to the left of $x$ since they are strictly less than $x$)

@@

If $x$ is even, then $x-1$ which is an odd number greater than both $y-1$ and $z-1$, must be to the left of $x$. Since $y-1$ and $z-1$ are already to the left of $x-1$, we conclude that swapping $y$ and $z$ still ensures that $y$ and $z$ are both to the right of $y-1$ and $z-1$.
\end{itemize}
Hence, we see that regardless of whether $x$ is even or odd, swapping $y$ and $z$ in a permutation $\pi \in T_n$ if $x > y > z$ or $x > z > y$ yields a permutation $\pi'$ that also belongs to $T_n$.
\end{proof}

Given these lemmas, we now present a proof for $t_n(312) = t_n(321)$.

\begin{theorem}
$t_n(312) = t_n(321)$
\end{theorem}

\begin{proof}
First observe from Lemma \ref{tn_avoidance} that if $x, y, z$ is a subsequence of three elements of a permutation in $T_n$ in the relative order $312$ or $321$, then both $y$ and $z$ must be even.

Before proving that there exists a bijective mapping between the set of permutations in $T_n$ that avoid $312$ (call this set $A$) and the set of permutations in $T_n$ that avoid $321$ (call this set $B$), we establish a chain of transformation moves that can be used to transform an element $a \in A$ to an element $b \in B$.

>>>>

\section{Avoidance of permutations of $T_n$}
\label{Tn}
We now impose further restrictions on the set of permutations $S_n$. This new set of permutations, $T_n$ is defined as follows.

\begin{definition}
Let $n$ be an even integer. The set $T_n$ is defined as the set of all permutations $\sigma \in S_n$ in which $1,3,5,\ldots,2n-1$ appear in increasing order, and $2i$ always appears to the right of $2i-1$.
\end{definition}

Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $0$.\todo{0 or 1 here?}

@@

\begin{proof}
Because $1,3,5,\ldots,2m-1$ must appear in increasing order in $T_{2m}$, the size of $T_{2m}$ is given by the number of ways in which $2, 4, 6, \ldots, 2m$ can be ordered relative to each other and to $1, 3, \ldots, 2m-1$.

Let us first insert $2m$ into the sequence $1,3,5,\ldots,2m-1$. Because $2m$ must appear to the right of $2m-1$, it can be inserted into only a single slot, the one following $2m-1$. (By ``slot'' we mean a gap between two existing elements of the sequence, or the gap that follows the last element of the sequence, or the gap that precedes the first element of the sequence.)

Next let us insert $(2m-2)$ into the sequence. $(2m-2)$ can be inserted into $3$ slots in the sequence -- the one between $2m-3$ and $2m-1$, the one between $2m-1$ and $2m$ or the one following $2m$.

We want to next insert $(2m-4)$ into this incompletely formed sequence; we see that there are $5$ possible locations. Continuing this for all even numbers up to $2$, we observe that the total number of ways such a sequence can  be created is equal to $1 \cdot 3 \cdot 5 \cdot \ldots \cdot (2m-1)$, as desired.
\end{proof}

Studying avoidance of permutations in $T_n$ seems an interesting exercise. In this section, we consider the sequence of numbers $t_n(\pi)$, defined as 
\[t_n(\pi)=\#\{\sigma\in T_n|\sigma \text{ avoids } \pi\},\]
with $\pi\in S_3$.

@@

More interesting are the equivalences between $213$ and $231$, and $312$ and $321$. We  prove the equivalence between $312$ and $321$ below, but before doing so, we first state and prove the following useful lemmas.

\begin{lemma}
\label{tn_avoidance}
Let $a$ and $b$ be two elements that appear in a permutation in $T_n$ such that $a$ is to the left of $b$. If $a > b$, then $b$ must be even.
\end{lemma}

\begin{proof}
For the sake of contradiction, let us assume that $b$ is odd. Then there exist two possibilities that need to be considered -- either $a$ is odd or $a$ is even.

@@

Note that all permutations in $T_n$ must have odd elements monotonically increasing. So if $a$ appears to the left of $b$, and both $a$ and $b$ are odd, then $a$ must be less than $b$ -- a contradiction. Hence $a$ cannot be odd.

@@

For all permutations $\sigma$ in $T_n$, every even element $j$ must be to the right of all odd elements less than $j$. Since $a$ is even, therefore, it must be to the right of all odd numbers less than it. But $b$ is an odd number less than $a$, and $a$ is to the left of $b$ -- another contradiction. Hence $a$ cannot be even, either.
\end{itemize}

We conclude that $b$ must be even, as desired.
\end{proof}

\begin{lemma}
\label{swapping_lemma}
If $x, y, z$ is a three-element subsequence of a permutation $\pi$ in $T_n$ such that $x > y > z$ or $x > z > y$, then swapping $y$ and $z$ in the permutation $\pi$ yields a permutation that also belongs to $T_n$.
\end{lemma}

\begin{proof}
Since $x$ is to the left of $y$ and $z$, and $x>y$ and $x>z$, it follows by Lemma \ref{tn_avoidance} that both $y$ and $z$ must be even. Let $\pi'$ be the permutation obtained by swapping $y$ and $z$ in the permutation $\pi$. We divide our analysis into two cases:
\begin{itemize}
\item \textbf{Case 1:} $x$ is odd.

Then it suffices to check that $y$ is to the right of $y-1$ and that $z$ is still to the right of $z-1$. This is easy: since $y-1$ and $z-1$ are odd numbers less than $x$ another odd number, they must be to the left of $x$, and hence to the left of $y$ and $z$.

@@

Since $\pi\in T_n$, $x-1$ must be to the left of $x$. Because $x>y$ and $x>z$, we must have that $x-1>y-1$ and $x-1>z-1$. It follows that $y-1$ and $z-1$ are to the left of $x-1$, and hence $x$, in $\pi$. We conclude that $y$ and $z$ are to the right of $y-1$ and $z-1$ in $\pi'$, since $y-1$ and $z-1$ are to the left of $x$ and $y$ and $z$ remain to the right of $x$.
\end{itemize}
Hence, regardless of the parity of $x$, swapping $y$ and $z$ yields a permutation $\pi'$ that also belongs to $T_n$, as desired.
\end{proof}

With these lemmas in hand, we now present a proof that $t_n(312) = t_n(321)$.

\begin{theorem}
$t_n(312)$, the set of permutations in $T_n$ avoiding 312, is precisely the same size as $t_n(321)$, the set of permutations in $T_n$ avoiding 321.
\end{theorem}

\begin{proof}
Let $A$ be the set of permutations in $T_n$ avoiding 312, and $B$ the set of permutations in $T_n$ avoiding 321. We will prove the theorem by constructing a bijection between $A$ and $B$.

Note by Lemma \ref{tn_avoidance} that if $x, y, z$ is a subsequence of three elements of a permutation in $T_n$ in the relative order $312$ or $321$, then both $y$ and $z$ must be even.

Before we do so, however, we establish a process to transform an element $a \in A$ to an element $b \in B$.
